XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-27 talk/icooolps2011/paper.tex
Typos and typographic details.

In this paper we present two of these hints that are extensively used in the
PyPy project to improve the performance of its Python interpreter. These
hints are used to control how the optimizer of the tracing JIT can improve the
traces of the object model. More specifically, these hints influence the
constant folding optimization. The first hint make it possible to turn arbitrary
variables in the trace into constants. The second hint allows the definition of
additional foldable operations.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project which used the for dynamic
assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions,
when a function call is encountered the operations of the called functions are
simply put into the trace too.

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used until the most important paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{XXX}.

@@

One disadvantage of tracing JITs which makes them not directly applicable to
PyPy is that they need to encode the language semantics of the language they are
tracing. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The numbers line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace indeed contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups somehow. How to achieve this will be topic of
Section~\ref{sec:fastobjmodel}.

@@

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993} there). Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} and by all older versions
of PyPy's JIT. Promotion is a technique that only works well in JIT compilers,
in static compilers it is significantly less applicable.

@@

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend to not vary often. Therefore it is possible to
promote the types. The next section will present a complete example of how
this works.

@@

In the last section we saw a way to turn arbitrary variables into constants. All
pure operations on these constants can be constant-folded. This works great for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not pure and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

@@

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

@@

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if the trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{write\_method} method on the class.

@@

SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
used to trace through an JavaScript implementation written in C\#. The
JavaScript implementation compiles JavaScript to CIL bytecode together with an
implementation of the JavaScript object model. The object model uses maps
and inline caches to speed up operations on objects. The tracer tracers through
the compiled JavaScript functions and the object model. SPUR contains two hints
that can be used to influence the tracer, one to prevent tracing of a C\#
function and one to force unrolling of a loop (PyPy has equivalent hints, but
they were not described in this paper).

>>>>

In this paper we present two of these hints that are extensively used in the
PyPy project to improve the performance of its Python interpreter. These
hints are used to control how the optimizer of the tracing JIT can improve the
traces of the object model. More specifically, these hints influence the
constant folding optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constants. The second hint allows the definition of
additional foldable operations.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project, which used one of them for dynamic
assembler optimization \cite{bala_dynamo:_2000}. Later they were used to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions:
when a function call is encountered the operations of the called functions are
simply put into the trace too.

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{XXX}.

@@

One disadvantage of (tracing) JITs which makes them not directly applicable to
PyPy is that they need to encode the language semantics of the language they are
tracing. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace indeed contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups somehow. How to achieve this will be topic of
Section~\ref{sec:fastobjmodel}.

@@

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993} there). Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} and by all older versions
of PyPy's JIT. Promotion is a technique that only works well in JIT compilers;
in static compilers it is significantly less applicable.

@@

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend not to vary often. Therefore it is possible to
promote the types. The next section will present a complete example of how
this works.

@@

In the previous section we saw a way to turn arbitrary variables into constants. All
pure operations on these constants can be constant-folded. This works great for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not pure and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

@@

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

XXX should we say ``version number'' all around when it is really an object?

@@

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if the trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{write\_method} method on the class.

@@

SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
used to trace through an JavaScript implementation written in C\#. The
JavaScript implementation compiles JavaScript to CIL bytecode together with an
implementation of the JavaScript object model. The object model uses maps
and inline caches to speed up operations on objects. The tracer traces through
the compiled JavaScript functions and the object model. SPUR contains two hints
that can be used to influence the tracer: one to prevent tracing of a C\#
function and one to force unrolling of a loop (PyPy has equivalent hints, but
they were not described in this paper).
